---
author: "Aaron Riedling"
output: html_document
---

## Introduction

In this assignment, you will work the dataset `gene_expression_data_wide`. The dataset contains gene expression measurements from two individuals, with 50 measurements per gene.
You will implement distance metrics, conduct hypothesis testing, and handle multiple testing scenarios.

## Task 1 (30 pts)
Create a function `calculate_euclidean` that computes the Euclidean distance between two gene expression vectors. Then:

a) Calculate the distance between Person1 and Person2's expression profiles (for each gene seperately)
b) Visualize the distances between both persons expression profiles

```{r task1_test}
# Your solution here!

# Load required libraries
library(ggplot2)

# Load the dataset
data <- read.csv("gene_expression_data_wide.csv")

# Function to calculate Euclidean distance
calculate_euclidean <- function(vec1, vec2) {
  sqrt(sum((vec1 - vec2)^2))
}

# Task 1a: Calculate distances for each gene
unique_genes <- unique(data$Gene)  # Identify unique genes
distances <- data.frame(
  Gene = character(),
  Euclidean_Distance = numeric()
)

for (gene in unique_genes) {
  gene_data <- subset(data, Gene == gene)
  distance <- calculate_euclidean(gene_data$Person1, gene_data$Person2)
  distances <- rbind(distances, data.frame(Gene = gene, Euclidean_Distance = distance))
}

# Task 1b: Visualize the distances
ggplot(distances, aes(x = Gene, y = Euclidean_Distance)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = "Euclidean Distances Between Person1 and Person2 for Each Gene",
       x = "Genes",
       y = "Euclidean Distance")



```

## Task 2 (35 pts)
Analyze the expression differences of GENE_1 between Person1 and Person2:

a) Conduct a t-test to compare expression levels between persons
b) Visualize the expression distributions using a [violin plot](https://en.wikipedia.org/wiki/Violin_plot)

```{r task2_test}
# Your solution here!

# Load necessary libraries
library(ggplot2)

# Task 2a: Conduct a t-test for GENE_1 between Person1 and Person2
# Subset the data for GENE_1
gene1_data <- subset(data, Gene == "GENE_1")

# Combine data into a long format for t-test and visualization
gene1_long <- data.frame(
  Expression = c(gene1_data$Person1, gene1_data$Person2),
  Person = rep(c("Person1", "Person2"), each = nrow(gene1_data))
)

# Perform a t-test
t_test_result <- t.test(Expression ~ Person, data = gene1_long)

# Print t-test results
print(t_test_result)

# Task 2b: Create a violin plot to visualize the expression distributions
ggplot(gene1_long, aes(x = Person, y = Expression, fill = Person)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, position = position_dodge(0.9), alpha = 0.9) + 
  labs(title = "Expression Distributions of GENE_1",
       x = "Person",
       y = "Expression Level") +
  theme_minimal() +
  theme(legend.position = "none")






```

## Task 3 (35 pts)
Perform a comprehensive analysis across all genes:

a) Conduct t-tests for all genes comparing Person1 vs Person2
b) Apply Bonferroni correction for multiple testing
c) Create a summary table with test statistics, p-values, and adjusted p-values
d) Visualize the results using a [Manhattan plot](https://en.wikipedia.org/wiki/Manhattan_plot)

```{r task3_test}
# Your solution here!
# Load necessary libraries
library(ggplot2)
library(dplyr)



# Task a: Conduct t-tests for all genes comparing Person1 vs Person2
# Initialize an empty data frame to store results
results <- data.frame(Gene = character(),
                      t_statistic = numeric(),
                      p_value = numeric(),
                      stringsAsFactors = FALSE)

# Loop through each gene
for (gene in unique(data$Gene)) {
  # Subset the data for the current gene
  gene_data <- subset(data, Gene == gene)
  
  # Conduct a t-test
  t_test <- t.test(gene_data$Person1, gene_data$Person2)
  
  # Store the results
  results <- rbind(results, data.frame(
    Gene = gene,
    t_statistic = t_test$statistic,
    p_value = t_test$p.value
  ))
}

# Task b: Apply Bonferroni correction for multiple testing
results$adjusted_p_value <- p.adjust(results$p_value, method = "bonferroni")

# Task c: Create a summary table
# Add a column indicating whether the test is significant after correction
results <- results %>%
  mutate(Significant = adjusted_p_value < 0.05)

# Print the summary table
print(results)

# Task d: Visualize the results using a Manhattan plot
# Add a negative log-transformed p-value column for plotting
results <- results %>%
  mutate(NegLogP = -log10(p_value))

# Create the Manhattan plot
ggplot(results, aes(x = Gene, y = NegLogP)) +
  geom_point(aes(color = Significant), size = 2) +
  scale_color_manual(values = c("FALSE" = "gray", "TRUE" = "red")) +
  labs(title = "Manhattan Plot of Gene Expression Differences",
       x = "Gene",
       y = "-log10(p-value)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))









```
